Towards a Verified Tableau Prover for a Quantifier-Free Fragment of Set Theory

### 2023-03-06 paper/CADE/paper.tex
paper: comments by Kevin

% Autoref
\usepackage[hidelinks]{hyperref}
\renewcommand{\sectionautorefname}{Section}
\renewcommand{\subsectionautorefname}{Section}

@@

\section{Introduction}
In Isabelle/HOL, there are specialised procedures for dealing with e.g.\ natural numbers, linear arithmetic, and metric spaces.
Some of these procedures have been verified in Isabelle/HOL such as a procedure for Presburger arithmetic~\cite{presburger} that was later extended to mixed real-integer arithmetic~\cite{arithmetic}.
This procedure, though, uses reflection to work on goals in Isabelle/HOL, which, during execution, either sacrifices speed by going through the simplifier or requires trusting the code generator.
More recently, \citeauthor{orders}~\cite{orders} presented a verified decision procedure for orders that produces certificates. 
This approach offers efficient execution by using generated code as well as soundness because the certificates are replayed through Isabelle's inference kernel.

@@

\subsection{Related Work}
Since the literature on decidable fragments of set theory is vast, we only focus on \MLSS{} here.
The fragment was first shown to be decidable by \citeauthor{mlss_first}~\cite{mlss_first}.
Subsequent work~\cite{mlss_np} found the decision problem to be \textbf{NP}-complete.
To obtain a practical decision procedure, \citeauthor{mlss_first_tableau}~\cite{mlss_first_tableau} proposed a tableau calculus, which was later improved by \citeauthor{tableau_quantifier_free}~\cite{tableau_quantifier_free}.
Both of these procedures construct a model during execution that is used to guide the proof search.
\citeauthor{tableau_quantifier_free} also cover an extension of the calculus with uninterpreted functions, which was revisited by \citeauthor{mlss_quantification}~\cite{mlss_quantification} who avoided the construction of a model during execution.
In this paper, we consider a version of the latter procedure due to \citeauthor{new_fast_tableau}~\cite{new_fast_tableau} that is specialised to \MLSS{} and where the branching rules of the calculus are set up in a way to guarantee mutual exclusivity of the branches.
Later extensions of the calculus added certain interpreted functions such as monotone functions~\cite{mlss_monotone_functions} and the inverse of a function~\cite{mlss_cartesian_map}.
The second extension notably includes the Cartesian product.
Those extensions, though, did not improve upon the tableau calculus for \MLSS{}.

@@

\section{Syntax and Semantics of MLSS\label{sec:semantics}}
\subsection{Syntax}
At the heart of \MLSS{}, we have the type of set terms which is the disjoint union of the empty set and variables as well as the operations union, intersection, set difference, and the singleton set represented by the constructor \lstinline!Single!.
We keep the type of variables abstract by making it a parameter of the set term data type.
The only restriction on the type of variables is that it needs to be infinite.
Isabelle/HOL's data type package automatically defines a function that gives us the set of variables in a set term, which we name \lstinline!vars!.
In what follows, we will overload the function \lstinline!vars! to also work on set atoms, formulae, and branches.
\begin{lstlisting}
datatype (vars: 'a) pset_term =
  $\emptyset$ | Var 'a | Single ('a pset_term)
| 'a pset_term $\unionS$ 'a pset_term | 'a pset_term $\interS$ 'a pset_term
| 'a pset_term $\diffS$ 'a pset_term
\end{lstlisting}
We can combine two set terms to form a set atom by using the membership or the equality operator.
\begin{lstlisting}
datatype (vars: 'a) pset_atom =
  'a pset_term $\inS$ 'a pset_term | 'a pset_term $\eqS$ 'a pset_term
\end{lstlisting}
With the above operators we can also represent the subset operator \lstinline!$\subseteqS$! and enumerate finite sets: \lstinline!s $\subseteqS$ t! is equivalent to \lstinline!s $\unionS$ t $\eqS$ t! and a finite set of elements \lstinline!{t$_\text{1}$,$\ldots$,t$_\text{k}$}! can be expressed by \lstinline!Single t$_1$ $\unionS$ $\ldots$ $\unionS$ Single t$_k$!.

@@

Likewise, we fix the constant \lstinline!subterms :: 'b $\Rightarrow$ 'a pset_term set!.
We overload this constant to return the set terms that are subterms of a set term, set atom, or formula, respectively. Lastly, we compute the subformulae of a formula with the function \lstinline!subfms :: 'a fm $\Rightarrow$ 'a fm set!.
The functions \lstinline!subterms! and \lstinline!subfms! are defined in the  way (cf.\ \appendixname~\ref{appendix:subterms}).

\subsection{Semantics}
The original paper~\cite{new_fast_tableau} bases the semantics of \MLSS{} on the von Neumann hierarchy of sets $\mathcal{V}$. 
While there is an entry~\cite{zfc_in_hol_afp} in the \emph{Archive of Formal Proofs} (AFP) that axiomatises this hierarchy,
we instead use the hierarchy of \emph{hereditarily finite sets} (HF sets) which fulfil all the same axioms as $\mathcal{V}$ --- that is, the axioms of ZF --- except for the axiom of infinity.
In particular, the membership relation is well-founded.
The HF sets are sufficient for our purposes as \MLSS{} is not powerful enough to force infinite models~\cite[Chapter~14.2]{set_theory}.
In contrast to $\mathcal{V}$, the HF sets are directly representable in Isabelle/HOL, and indeed, there is an AFP entry~\cite{hf_AFP} that formalises them.
The entry defines a type \lstinline!hf! that comes with the following functionality:
\begin{itemize}
  \item The function \lstinline!HF :: hf set $\Rightarrow$ set! that converts a finite set of HF sets into an HF set.
\item The usual set operations such as equality ($=$), membership ($\hfmem$), union ($\sqcup$), intersection ($\sqcap$), and difference ($-$) are defined.
\item Finally, the empty set coincides with the ordinal $0$, so it is denoted by \lstinline!0 :: hf!.
\end{itemize}

@@

\subsection{Linear Expansion Rules}
\begin{table*}[t]
  \centering
  \lstset{
    escapeinside={@}{@}
  }
  \newcommand{\ton}{\text{t}_\text{1}}
  \newcommand{\ttw}{\text{t}_\text{2}}
  \caption{Linear expansion rules.\label{tab:lexpands} We omit the rules for $\interS$ and $\diffS$ as they coincide with the original paper (cf.\ \appendixname~\ref{appendix:linear_rules}).}
  \begin{tabular}{ccc}
    \toprule
    Propositional Rules & {\hskip 6pt} & Rules for \lstinline!$\unionS$! \\
    \cmidrule(l{1pt}){1-1}\cmidrule(r{1pt}){3-3}
    \begin{tabular}{p{0.18\textwidth}m{0.05\textwidth}p{0.175\textwidth}}
      \lstinline!@\fmAnd{p}{q}@! & $\Longrightarrow$ & \lstinline!p, q! \\
      \lstinline!@\fmNeg{(\fmOr{p}{q})}@! & $\Longrightarrow$ & \lstinline!@\fmNeg{p}@, @\fmNeg{q}@! \\
      \lstinline!@\fmOr{p}{q}, \fmNeg{p}@! & $\Longrightarrow$ & \lstinline!q! \\
      \lstinline!@\fmOr{p}{q}, \fmNeg{q}@! & $\Longrightarrow$ & \lstinline!p! \\
      \lstinline!@\fmNeg{(\fmAnd{p}{q})}@, p! & $\Longrightarrow$ & \lstinline!@\fmNeg{q}@! \\
      \lstinline!@\fmNeg{(\fmAnd{p}{q})}@, q! & $\Longrightarrow$ & \lstinline!@\fmNeg{p}@! \\
      \lstinline!@\fmNeg{(\fmNeg{p})}@! & $\Longrightarrow$ & \lstinline!p!\\[6ex]
    \end{tabular}
                        & &
    \begin{tabular}{p{0.22\textwidth}m{0.05\textwidth}p{0.22\textwidth}}
    \lstinline!s $\notinS$ $\ton$ $\unionS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\notinS$ $\ton$, s $\notinS$ $\ttw$! \\
    \lstinline!s $\inS$ $\ton$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ton$ $\unionS$ $\ttw$! \\
    \lstinline!s $\in$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ton$ $\unionS$ $\ttw$! \\
    \lstinline!s $\inS$ $\ton$ $\unionS$ $\ttw$!, \newline \lstinline!s $\notinS$ $\ton$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ttw$! \\
    \lstinline!s $\inS$ $\ton$ $\unionS$ $\ttw$!, \newline \lstinline!s $\notinS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ton$! \\
    \lstinline!s $\notinS$ $\ton$, s $\notinS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\notinS$ $\ton$ $\unionS$ $\ttw$! \\[0.9em]
    \end{tabular}
    \\

@@

The calculus considers two kinds of branch expansion rules: \emph{linear} and \emph{branching} rules.
As the name suggests, branching rules lead to the creation of new branches in the tableau while linear rules only extend a branch \lstinline!b! with new formulae \lstinline!b' = [$\psi_1$,$\ldots$,$\psi_n$]!, which we denote by \lstinline!&\lexpands{b'}{b}&!.
\autoref{tab:lexpands} shows the linear expansion rules.
Note that in the first two rules for \lstinline!$\eqS$!, \lstinline!l! is a literal occurring in the branch.
Furthermore, the term-for-term substitution \lstinline!l{s/t}! is restricted to replace \lstinline!t! only if it occurs as a top-level set terms of \lstinline!l!.
As an example, \lstinline!u! and \lstinline!s $\interS$ t! are top-level terms of \lstinline!u $\inS$ s $\interS$ t! but \lstinline!s! and \lstinline!t! are not.

The more crucial restriction of the linear rules is that that no new subterm may be created by their application;
for instance, the second rule for \lstinline!$\unionS$! is
\lstinline!s $\inS$ t$_\text{1}$ $\Longrightarrow$ s $\inS$ t$_\text{1}$ $\unionS$ t$_\text{2}$!,
which formally represents 
\begin{lstlisting}
  (s $\inS$ t$_\text{1}$) $\in$ set b $\Longrightarrow$ &\lexpands{[s $\inS$ t$_\text{1}$ $\unionS$ t$_\text{2}$]}{b}&,
\end{lstlisting}
may only be used under the condition
\lstinline[breaklines]!t$_\text{1}$ $\unionS$ t$_\text{2}$ $\in$ subterms (last b)!.
The purpose of this restriction is to prevent unbounded expansion of the branch.
In fact, we give an explicit upper bound for the number of formulae in a branch in \autoref{sec:correct}.

@@

\subsection{Branching Rules\label{sec:branching}}
\begin{table*}[t]
  \lstset{
    escapeinside={@}{@}
  }
  \centering
  \newcommand{\ton}{\text{t}_\text{1}}
  \newcommand{\ttw}{\text{t}_\text{2}}
  \newcommand{\lpipe}{\rule[-0.4ex]{0.6pt}{2.3ex}}
  \caption{Branching expansion rules.\label{tab:bexpands} We write \lstinline!$\phi$! for \lstinline!last b! here. The omitted rules coincide with the original paper (cf.\ \appendixname~\ref{appendix:branching_rules}).}
  \begin{tabular}{>{\centering\arraybackslash}p{0.34\textwidth} >{\centering\arraybackslash}p{0.32\textwidth} >{\centering\arraybackslash}p{0.3\textwidth}}
    \toprule
    Rule & Precondition & Subsumption condition \\
    \midrule
    \begin{tabular}{@{}>{\raggedleft\arraybackslash}p{0.11\textwidth}c>{\raggedright\arraybackslash}p{0.11\textwidth}@{}}
      \\
      \midrule
      \lstinline!p! & \lpipe & \lstinline!@\fmNeg{p}@! \\
    \end{tabular}
         & \lstinline!@\fmOr{p}{q}@ $\in$ set b! &
         \begin{tabular}{rl}
         & \lstinline!p $\in$ set b $\lor$! \\
         $\fmNegSymbol$ & \lstinline!p $\in$ set b! \\
         \end{tabular} \\

@@

           \lstinline!$\exists$s.! & \lstinline!(s $\inS$ $\ton$) $\in$ set b! \\
               \lstinline!$\land$! & \lstinline!(s $\notinS$ $\ttw$) $\in$ set b! \\
           \multicolumn{2}{c}{$\lor$} \\
           \lstinline!$\exists$s.! & \lstinline!(s $\notinS$ $\ton$) $\in$ set b! \\
               \lstinline!$\land$! & \lstinline!(s $\inS$ $\ttw$) $\in$ set b! \\
         \end{tabular}
         \\
         \bottomrule
  \end{tabular}
\end{table*}
After running out of linear rules to apply, only the branching rules shown in \autoref{tab:bexpands} remain.
A rule is applicable if its \emph{precondition} is met and, to prevent unnecessary branching, it is not subsumed as indicated by the \emph{subsumption condition}.
These rules create multiple branches in the tableau, so we represent the different possibilities \lstinline!bs'! to expand a branch \lstinline!b! as a set and write \lstinline!&\fexpands{bs'}{b}&!. 
Accordingly, we get a new branch \lstinline!b' @ b! in the tableau for each \lstinline!b' $\in$ set bs'!.

@@

We only discuss the abstract specification here and refer the reader to the formalisation for the executable specification.
The implementation uses a couple of features of Isabelle/HOL's function package:
instead of defining the function via pattern matching, we specify the equations of the function as conditional rewrite rules.
This requires us to prove that the assumptions of the equations are non-overlapping, which is done by automation.
The other concern is that Isabelle/HOL requires functions to be total, so a recursive function needs to terminate in order for it to be well-defined;
nevertheless, the termination proof is separated from the definition of the function for modularity.
The function package maintains the soundness of the definition by introducing a so-called domain predicate \lstinline!mlss_proc_branch_dom! which characterises the arguments for which the function terminates.
Each equation of the function is guarded by an assumption that the predicate holds for the argument.
In \autoref{sec:correct}, we will show that the domain predicate holds for the context that \lstinline!mlss_proc_branch! is called in.
Before we go into more detail on how the termination is proved, we discuss its definition as shown below.

@@

In particular, we will use this notion in \autoref{sec:correct} to state an upper bound for the cardinality of well-formed branches, with which we justify the termination of the decision procedure. 
Before we come to that, though, we prove soundness and completeness in Section~\ref{sec:sound} and~\ref{sec:complete}, respectively.
In \autoref{sec:correct}, we also show that both properties easily transfer to \lstinline!mlss_proc! which, together with termination, establishes that it is a decision procedure.

\section{Completeness of the Calculus\label{sec:complete}}
For completeness of the calculus, we need to show that every unsatisfiable formula has a closed tableau or, conversely, that the formula is satisfiable if there is a saturated and open branch in the tableau.
To facilitate inductive reasoning, we show a stronger statement by constructing a model \lstinline!M! such that \lstinline!M $\models$ $\phi$! for all \lstinline!$\phi$ $\in$ set b!.
At the core of the model, there is a \emph{realisation} function that maps set terms to sets of type \lstinline!hf!.
A subset of the witnesses, which we call \emph{pure} witnesses, receives special treatment from the realisation function for reasons that will become apparent in \autoref{sec:pwits}.
The collection of set terms of a branch can thus be partitioned into two collections as defined below.
\begin{lstlisting}
definition pwits :: 'a branch $\Rightarrow$ 'a set where
  pwits b $\equiv$ {c $\in$ wits b. $\forall$t $\in$ subterms (last b).
         $\phantom{\equiv}$     AT (Var c $\eqS$ t) $\notin$ set b $\land$ AT (t $\eqS$ Var c) $\notin$ set b}

@@

Before we prove that the realisation function constitutes a model in \autoref{sec:real}, we will first explain the significance of the pure witnesses.

\subsection{Characterisation of the Pure Witnesses\label{sec:pwits}}
Recall that the pure witnesses of a branch \lstinline!b! are those witnesses that are not related to other subterms in \lstinline!last b! by equality.
In the context of a well-formed branch, this characterisation can be strengthened to any set term and, in addition, we also get that there is no membership literal where a pure witness is on the right-hand side.
Intuitively speaking, the realisation of a pure witness does not depend on the realisation of any other set term.
\begin{lstlisting}[label={lst:lemma_2}]
lemma lemma_2:
  assumes wf_branch b and c $\in$ pwits b
  shows (Var c $\eqS$ t) $\notin$ set b and (t $\eqS$ Var c) $\notin$ set b
    and (t $\inS$ Var c) $\notin$ set b
\end{lstlisting}
\newcommand{\reflemmatwo}{\hyperref[lst:lemma_2]{\lstinline!lemma_2!}}
So why are pure witnesses treated differently?
According to the definition of \lstinline!realise!, the pure witnesses would be evaluated to the empty set \lstinline!0 :: hf!, were they not treated separately. 
To see that this is a problem, consider the branch
\lstinline!b = [Var s $\neqS$ Var t, Var t $\neqS$ Var u]!
which expands to several open and saturated branches, one of which is
\begin{lstlisting}
  [Var x $\neqS$ Var y, Var x $\inS$ Var s, Var x $\notinS$ Var t,
  &\phantom{[Var x $\neqS$ Var y,}& Var y $\inS$ Var t, Var y $\notinS$ Var u] @ b
\end{lstlisting}
for some fresh \lstinline!x! and \lstinline!y!.
Assigning both \lstinline!Var x! and \lstinline!Var y! a value of \lstinline!0! would contradict the literal \lstinline!Var x $\neqS$ Var y!.
To prevent this, we assign the pure witnesses pairwise different values.

>>>>

% Autoref
\usepackage[hidelinks]{hyperref}
\usepackage[capitalise]{cleveref}

@@

\section{Introduction}
In Isabelle/HOL, there are specialised procedures for dealing with e.g.\ natural numbers, linear arithmetic, and metric spaces.
Some of these procedures have been verified in Isabelle/HOL, such as a procedure for Presburger arithmetic~\cite{presburger} that was later extended to mixed real-integer arithmetic~\cite{arithmetic}.
This procedure, though, uses reflection to work on goals in Isabelle/HOL, which, during execution, either sacrifices speed by going through the simplifier or requires trusting the code generator.
More recently, \citeauthor{orders}~\cite{orders} presented a verified decision procedure for orders that produces certificates. 
This approach offers efficient execution by using generated code as well as soundness because the certificates are replayed through Isabelle's inference kernel.

@@

\subsection{Related Work}
Since the literature on decidable fragments of set theory is vast, we only focus on \MLSS{} here.
The fragment was first shown to be decidable by \citeauthor{mlss_first}~\cite{mlss_first}.
Subsequent work~\cite{mlss_np} found the decision problem to be \textbf{NP}-complete.
To obtain a practical decision procedure, \citeauthor{mlss_first_tableau}~\cite{mlss_first_tableau} proposed a tableau calculus, which was later improved by \citeauthor{tableau_quantifier_free}~\cite{tableau_quantifier_free}.
Both of these procedures construct a model during execution that is used to guide the proof search.
\citeauthor{tableau_quantifier_free} also cover an extension of the calculus with uninterpreted functions, which was revisited by \citeauthor{mlss_quantification}~\cite{mlss_quantification} who avoided the construction of a model during execution.
In this paper, we consider a version of the latter procedure due to \citeauthor{new_fast_tableau}~\cite{new_fast_tableau} that is specialised to \MLSS{} and where the branching rules of the calculus are set up in a way to guarantee mutual exclusivity of the branches.
Later extensions of the calculus added certain interpreted functions such as monotone functions~\cite{mlss_monotone_functions} and the inverse of a function~\cite{mlss_cartesian_map}.
The latter extension notably includes the Cartesian product.
Those extensions, though, did not improve upon the tableau calculus for \MLSS{}.

@@

\section{Syntax and Semantics of MLSS\label[section]{sec:semantics}}
\subsection{Syntax}
At the heart of \MLSS{}, we have the type of set terms which is the disjoint union of the empty set and variables as well as the operations union, intersection, set difference, and the singleton set represented by the constructor \lstinline!Single!.
We keep the type of variables abstract by making it a parameter of the set term data type.
The only restriction on the type of variables is that it needs to be infinite.
Isabelle/HOL's data type package automatically defines a function that gives us the set of variables in a set term, which we name \lstinline!vars!.
In what follows, we will overload the function \lstinline!vars! to also work on set atoms, formulae, and branches.
\begin{lstlisting}
datatype (vars: 'a) pset_term =
  $\emptyset$ | Var 'a | Single ('a pset_term)
| 'a pset_term $\unionS$ 'a pset_term | 'a pset_term $\interS$ 'a pset_term
| 'a pset_term $\diffS$ 'a pset_term
\end{lstlisting}
We can combine two set terms to form a set atom by using the membership or the equality operator.
\begin{lstlisting}
datatype (vars: 'a) pset_atom =
  'a pset_term $\inS$ 'a pset_term | 'a pset_term $\eqS$ 'a pset_term
\end{lstlisting}
With the above operators we can also represent the subset operator \lstinline!$\subseteqS$! and enumerate finite sets: \lstinline!s $\subseteqS$ t! is equivalent to \lstinline!s $\unionS$ t $\eqS$ t! and a finite set of elements \lstinline!{t$_\text{1}$,$\ldots$,t$_\text{k}$}! can be expressed by \lstinline!Single t$_1$ $\unionS$ $\ldots$ $\unionS$ Single t$_k$!.

@@

Likewise, we fix the constant \lstinline!subterms :: 'b $\Rightarrow$ 'a pset_term set!.
We overload this constant to return the set terms that are subterms of a set term, set atom, or formula, respectively. Lastly, we compute the subformulae of a formula with the function \lstinline!subfms :: 'a fm $\Rightarrow$ 'a fm set!.
The functions \lstinline!subterms! and \lstinline!subfms! are defined in the expected way (cf.\ \appendixname~\ref{appendix:subterms}).

\subsection{Semantics}
The original paper~\cite{new_fast_tableau} bases the semantics of \MLSS{} on the von Neumann hierarchy of sets $\mathcal{V}$. 
While there is an entry~\cite{zfc_in_hol_afp} in the \emph{Archive of Formal Proofs} (AFP) that axiomatises this hierarchy,
we instead use the hierarchy of \emph{hereditarily finite sets} (HF sets) which fulfil all the same axioms as $\mathcal{V}$ -- that is, the axioms of ZF -- except for the axiom of infinity.
In particular, the membership relation is well-founded.
The HF sets are sufficient for our purposes as \MLSS{} is not powerful enough to force infinite models~\cite[Chapter~14.2]{set_theory}.
In contrast to $\mathcal{V}$, the HF sets are directly representable in Isabelle/HOL, and indeed, there is an AFP entry~\cite{hf_AFP} that formalises them.
The entry defines a type \lstinline!hf! that comes with the following functionality:
\begin{itemize}
  \item The function \lstinline!HF :: hf set $\Rightarrow$ set! that converts a finite set of HF sets into an HF set.
\item The usual set operations such as equality ($=$), membership ($\hfmem$), union ($\sqcup$), intersection ($\sqcap$), and difference ($-$) are defined.
\item Finally, the empty set coincides with the ordinal $0$, so it is denoted by \lstinline!0 :: hf!.
\end{itemize}

@@

\subsection{Linear Expansion Rules}
\begin{table*}[t]
  \centering
  \lstset{
    escapeinside={@}{@}
  }
  \newcommand{\ton}{\text{t}_\text{1}}
  \newcommand{\ttw}{\text{t}_\text{2}}
  \caption{Linear expansion rules.\label[table]{tab:lexpands} We omit the rules for $\interS$ and $\diffS$ as they coincide with the original paper (cf.\ \appendixname~\ref{appendix:linear_rules}).}
  \begin{tabular}{ccc}
    \toprule
    Propositional Rules & {\hskip 6pt} & Rules for \lstinline!$\unionS$! \\
    \cmidrule(l{1pt}){1-1}\cmidrule(r{1pt}){3-3}
    \begin{tabular}{p{0.18\textwidth}m{0.05\textwidth}p{0.175\textwidth}}
      \lstinline!@\fmAnd{p}{q}@! & $\Longrightarrow$ & \lstinline!p, q! \\
      \lstinline!@\fmNeg{(\fmOr{p}{q})}@! & $\Longrightarrow$ & \lstinline!@\fmNeg{p}@, @\fmNeg{q}@! \\
      \lstinline!@\fmOr{p}{q}, \fmNeg{p}@! & $\Longrightarrow$ & \lstinline!q! \\
      \lstinline!@\fmOr{p}{q}, \fmNeg{q}@! & $\Longrightarrow$ & \lstinline!p! \\
      \lstinline!@\fmNeg{(\fmAnd{p}{q})}@, p! & $\Longrightarrow$ & \lstinline!@\fmNeg{q}@! \\
      \lstinline!@\fmNeg{(\fmAnd{p}{q})}@, q! & $\Longrightarrow$ & \lstinline!@\fmNeg{p}@! \\
      \lstinline!@\fmNeg{(\fmNeg{p})}@! & $\Longrightarrow$ & \lstinline!p!\\[6ex]
    \end{tabular}
                        & &
    \begin{tabular}{p{0.22\textwidth}m{0.05\textwidth}p{0.22\textwidth}}
    \lstinline!s $\notinS$ $\ton$ $\unionS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\notinS$ $\ton$, s $\notinS$ $\ttw$! \\
    \lstinline!s $\inS$ $\ton$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ton$ $\unionS$ $\ttw$! \\
    \lstinline!s $\in$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ton$ $\unionS$ $\ttw$! \\
    \lstinline!s $\inS$ $\ton$ $\unionS$ $\ttw$!, \newline \lstinline!s $\notinS$ $\ton$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ttw$! \\
    \lstinline!s $\inS$ $\ton$ $\unionS$ $\ttw$!, \newline \lstinline!s $\notinS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ton$! \\
    \lstinline!s $\notinS$ $\ton$, s $\notinS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\notinS$ $\ton$ $\unionS$ $\ttw$! \\[0.9em]
    \end{tabular}
    \\

@@

The calculus considers two kinds of branch expansion rules: \emph{linear} and \emph{branching} rules.
As the name suggests, branching rules lead to the creation of new branches in the tableau while linear rules only extend a branch \lstinline!b! with new formulae \lstinline!b' = [$\psi_1$,$\ldots$,$\psi_n$]!, which we denote by \lstinline!&\lexpands{b'}{b}&!.
\cref{tab:lexpands} shows the linear expansion rules.
Note that in the first two rules for \lstinline!$\eqS$!, \lstinline!l! is a literal occurring in the branch.
Furthermore, the term-for-term substitution \lstinline!l{s/t}! is restricted to replace \lstinline!t! only if it occurs as a top-level set term of \lstinline!l!.
As an example, \lstinline!u! and \lstinline!s $\interS$ t! are top-level terms of \lstinline!u $\inS$ s $\interS$ t! but \lstinline!s! and \lstinline!t! are not.

The more crucial restriction of the linear rules is that that no new subterm may be created by their application;
for instance, the second rule for \lstinline!$\unionS$! is
\lstinline!s $\inS$ t$_\text{1}$ $\Longrightarrow$ s $\inS$ t$_\text{1}$ $\unionS$ t$_\text{2}$!,
which formally represents 
\begin{lstlisting}
  (s $\inS$ t$_\text{1}$) $\in$ set b $\Longrightarrow$ &\lexpands{[s $\inS$ t$_\text{1}$ $\unionS$ t$_\text{2}$]}{b}&,
\end{lstlisting}
may only be used under the condition
\lstinline[breaklines]!t$_\text{1}$ $\unionS$ t$_\text{2}$ $\in$ subterms (last b)!.
The purpose of this restriction is to prevent unbounded expansion of the branch.
In fact, we give an explicit upper bound for the number of formulae in a branch in \cref{sec:correct}.

@@

\subsection{Branching Rules\label[section]{sec:branching}}
\begin{table*}[t]
  \lstset{
    escapeinside={@}{@}
  }
  \centering
  \newcommand{\ton}{\text{t}_\text{1}}
  \newcommand{\ttw}{\text{t}_\text{2}}
  \newcommand{\lpipe}{\rule[-0.4ex]{0.6pt}{2.3ex}}
  \caption{Branching expansion rules.\label[table]{tab:bexpands} We write \lstinline!$\phi$! for \lstinline!last b! here. The elided rules coincide with the original paper (cf.\ \appendixname~\ref{appendix:branching_rules}).}
  \begin{tabular}{>{\centering\arraybackslash}p{0.34\textwidth} >{\centering\arraybackslash}p{0.32\textwidth} >{\centering\arraybackslash}p{0.3\textwidth}}
    \toprule
    Rule & Precondition & Subsumption condition \\
    \midrule
    \begin{tabular}{@{}>{\raggedleft\arraybackslash}p{0.11\textwidth}c>{\raggedright\arraybackslash}p{0.11\textwidth}@{}}
      \\
      \midrule
      \lstinline!p! & \lpipe & \lstinline!@\fmNeg{p}@! \\
    \end{tabular}
         & \lstinline!@\fmOr{p}{q}@ $\in$ set b! &
         \begin{tabular}{rl}
         & \lstinline!p $\in$ set b $\lor$! \\
         $\fmNegSymbol$ & \lstinline!p $\in$ set b! \\
         \end{tabular} \\

@@

           \lstinline!$\exists$s.! & \lstinline!(s $\inS$ $\ton$) $\in$ set b! \\
               \lstinline!$\land$! & \lstinline!(s $\notinS$ $\ttw$) $\in$ set b! \\
           \multicolumn{2}{c}{$\lor$} \\
           \lstinline!$\exists$s.! & \lstinline!(s $\notinS$ $\ton$) $\in$ set b! \\
               \lstinline!$\land$! & \lstinline!(s $\inS$ $\ttw$) $\in$ set b! \\
         \end{tabular}
         \\
         \bottomrule
  \end{tabular}
\end{table*}
After running out of linear rules to apply, only the branching rules shown in \cref{tab:bexpands} remain.
A rule is applicable if its \emph{precondition} is met and, to prevent unnecessary branching, it is not subsumed as indicated by the \emph{subsumption condition}.
These rules create multiple branches in the tableau, so we represent the different possibilities \lstinline!bs'! to expand a branch \lstinline!b! as a set and write \lstinline!&\fexpands{bs'}{b}&!. 
Accordingly, we get a new branch \lstinline!b' @ b! in the tableau for each \lstinline!b' $\in$ set bs'!.

@@

We only discuss the abstract specification here and refer the reader to the formalisation for the executable specification.
The implementation uses a couple of features of Isabelle/HOL's function package:
instead of defining the function via pattern matching, we specify the equations of the function as conditional rewrite rules.
This requires us to prove that the assumptions of the equations are non-overlapping, which is done by automation.
The other concern is that Isabelle/HOL requires functions to be total, so a recursive function needs to terminate in order for it to be well-defined;
nevertheless, the termination proof is separated from the definition of the function for modularity.
The function package maintains the soundness of the definition by introducing a so-called domain predicate \lstinline!mlss_proc_branch_dom! which characterises the arguments for which the function terminates.
Each equation of the function is guarded by an assumption that the predicate holds for the argument.
In \cref{sec:correct}, we will show that the domain predicate holds for the context that \lstinline!mlss_proc_branch! is called in.
Before we go into more detail on how the termination is proved, we discuss its definition as shown below.

@@

In particular, we use this notion in \cref{sec:correct} to state an upper bound for the cardinality of well-formed branches.
The upper bound justifies the termination of the decision procedure. 
Before we come to that, though, we prove soundness and completeness in Section~\ref{sec:sound} and~\ref{sec:complete}, respectively.
In \cref{sec:correct}, we also show that both properties easily transfer to \lstinline!mlss_proc! which, together with termination, establishes that it is a decision procedure.

\section{Completeness of the Calculus\label[section]{sec:complete}}
For completeness of the calculus, we need to show that every unsatisfiable formula has a closed tableau or, conversely, that the formula is satisfiable if there is a saturated and open branch in the tableau.
To facilitate inductive reasoning, we show a stronger statement by constructing a model \lstinline!M! such that \lstinline!M $\models$ $\phi$! for all \lstinline!$\phi$ $\in$ set b!.
At the core of the model, there is a \emph{realisation} function that maps set terms to sets of type \lstinline!hf!.
A subset of the witnesses, which we call \emph{pure} witnesses, receives special treatment from the realisation function for reasons that will become apparent in \cref{sec:pwits}.
The collection of set terms of a branch can thus be partitioned into two collections as defined below.
\begin{lstlisting}
definition pwits :: 'a branch $\Rightarrow$ 'a set where
  pwits b $\equiv$ {c $\in$ wits b. $\forall$t $\in$ subterms (last b).
         $\phantom{\equiv}$     AT (Var c $\eqS$ t) $\notin$ set b $\land$ AT (t $\eqS$ Var c) $\notin$ set b}

@@

Before we prove that the realisation function constitutes a model in \cref{sec:real}, we will first explain the significance of the pure witnesses.

\subsection{Characterisation of the Pure Witnesses\label[section]{sec:pwits}}
Recall that the pure witnesses of a branch \lstinline!b! are those witnesses that are not related to other subterms in \lstinline!last b! by equality.
In the context of a well-formed branch, this characterisation can be strengthened to any set term and, in addition, we also get that there is no membership literal where a pure witness is on the right-hand side.
Intuitively speaking, the realisation of a pure witness does not depend on the realisation of any other set term.
\begin{lstlisting}[label={lst:lemma_2}]
lemma lemma_2:
  assumes wf_branch b and c $\in$ pwits b
  shows (Var c $\eqS$ t) $\notin$ set b and (t $\eqS$ Var c) $\notin$ set b
    and (t $\inS$ Var c) $\notin$ set b
\end{lstlisting}
\newcommand{\reflemmatwo}{\hyperref[lst:lemma_2]{\lstinline!lemma_2!}}
So why are pure witnesses treated differently?
According to the definition of \lstinline!realise!, the pure witnesses would be evaluated to the empty set \lstinline!0 :: hf!, were they not treated separately. 
To see that this is a problem, consider the branch
\lstinline!b = [Var s $\neqS$ Var t, Var t $\neqS$ Var u]!
which expands to several open and saturated branches, one of which is
\begin{lstlisting}
  [Var x $\neqS$ Var y, Var x $\inS$ Var s, Var x $\notinS$ Var t,
  &\phantom{[Var x $\neqS$ Var y,}& Var y $\inS$ Var t, Var y $\notinS$ Var u] @ b
\end{lstlisting}
for some fresh \lstinline!x! and \lstinline!y!.
Assigning both \lstinline!Var x! and \lstinline!Var y! a value of \lstinline!0! would contradict the literal \lstinline!Var x $\neqS$ Var y!.
To prevent this, we assign the pure witnesses pairwise different values.
